---
title: "AM14 Assignment 1 - Sammy"
author: "Sammy Chen"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing Packages

```{r load_libraries}
library(readxl)
library(tidyverse)
library(lubridate)
library(PerformanceAnalytics)
library(ggthemes)

```

> 1. Go to CANVAS and download the data for PS1 Monthly.xlsx. The data was downloaded from the CRSP (The Center for Research in Security Prices) data base via WRDS (Wharton Research Service) that is available for LBS students. Note the file contains data about Microsoft, Exxon Mobil (previously Exxon), General Electric, JP Morgan Chase (previously Chemical Banking and Chase Manhattan), Intel, Citigroup (previously Primerica and Travellers Group). In addition, the columns vwretd (ewretd) and vwredx (ewretx) contain value-weighted (equal-weighted) total returns and total returns excluding dividends for the CRSP index that contains stocks from NYSE, AMEX, and NASDAQ. Finally, sprtrn contains the total return for the S&P 500 Composite Index.

# Loading in the Data

```{r loading_data}

#loading in PS1 monthly data
PS1_Monthly <- read_excel("PS1_Monthly.xlsx") %>% 
  mutate(DATE = lubridate::ymd(date)) %>% #fix `date`
  select(1,2,19,3:18)

#addressing missingness in DIVAMT by imputing with 0
PS1_Monthly[is.na(PS1_Monthly)] <- 0

```

> 2. Make sure you understand how the holding period returns (RET) are calculated given the (unadjusted) prices (PRC), dividends (DIVAMT) and adjustment for the number of shares (CFACPR, i.e., the adjusted price equals PRC/CFACPR). That is, replicate the returns using the raw data. You can find more information about the variables on the CRSP page online. Calculate returns where you omit the dividends, i.e., you focus solely on capital gains (call these returns RETX).

```{r creating_testing_dataframe}

#creating a testing data frame
testing <- PS1_Monthly %>% 
  select(3,4,9,7,13,10,14)

```

**The Calculation of Holding Period Returns (RET) is as follows:**

$RET_t \ = \ \frac{(PRC_t + DIVAMT_t)/CFACPR_t}{PRC_{t-1}/CFACPR_{t-1}}-1$

```{r calculating_RET}

#calculating RET from PRC, DIVAMT and CFACPR
testing <- testing %>% 
  group_by(TICKER) %>% 
  mutate(RET_rep = round(((((PRC+DIVAMT)/CFACPR)/(lag(PRC)/lag(CFACPR)))-1),6)) %>% 
  ungroup()
           
#evaluating whether this calculation was successful
summary(testing$RET==testing$RET_rep)


```

**The Calculation of Holding Period Returns, Omitting Dividends (RETX) is as follows:**

$RETX_t \ = \ \frac{(PRC_t)/CFACPR_t}{PRC_{t-1}/CFACPR_{t-1}}-1$

```{r calculating_RETX}


#calculating RETX
testing <- testing %>% 
  group_by(TICKER) %>% 
  mutate(RETX_rep = round(((((PRC)/CFACPR)/(lag(PRC)/lag(CFACPR)))-1),6)) %>% 
  ungroup()
  
#evaluating whether this calculation was successful
summary(testing$RETX==testing$RETX_rep)


```

> 3. Use the holding period returns to create a total return index for the MSFT and GE stocks and the S&P 500 index, which shows the theoretical growth in value of an investment in the stock assuming that dividends are reinvested (normalize the start value to 1). Do the same for the returns that abstract from dividend payments (i.e., use RETX instead). Plot the investments with and without dividends for each stock separately. How do dividends affect the results stock by stock?

```{r}

PS1_Monthly <- PS1_Monthly %>% 
  group_by(TICKER) %>% 
#creating a total return index for the MSFT and GE stocks and the S&P 500 index assuming dividends are reinvested (normalizing the start value to 1)
  mutate(total_return_index_div = RET/RET[2]) %>%  
  mutate(total_return_index_no_div = RETX/RETX[2]) %>% 
  mutate(total_return_index_SP500 = sprtrn/sprtrn[1])

MSFT_return_div <- PS1_Monthly %>% 
  filter(TICKER == "MSFT") %>% 
  ggplot(aes(x = total_return_index_div,
             y = total_return_index_no_div)) +
    geom_point()+
    labs(title = "Investment with dividends - MSFT", 
       x = "total return index with dividends", 
       y = "total return index without dividends")+
    theme_fivethirtyeight()+
    NULL

MSFT_return_div

GE_return_div <- PS1_Monthly %>% 
  filter(TICKER == "GE") %>% 
  ggplot(aes(x = total_return_index_div,
             y = total_return_index_no_div)) +
  geom_point()+
  labs(title = "Investment with dividends - GE", 
       x = "total return index with dividends", 
       y = "total return index without dividends")+
    theme_fivethirtyeight()+
    NULL

GE_return_div

```

> 4.  Normalize the price of GE using the adjusted number of shares. Plot the adjusted price against the unadjusted prices. Discuss.

```{r}

PS1_Monthly %>% 
  filter(TICKER == "GE") %>% 
  mutate(adjusted_price = PRC/CFACPR) %>% #normailize the price by adjusting for number of shares
  ggplot(aes(x = PRC, y = adjusted_price))+
  geom_point()+
  labs(title = "Comparison of GE prices", 
       x = "Unadjusted price", 
       y = "Adjusted price")+
  theme_fivethirtyeight()+
    NULL

```

> 5.  The holding period returns are normal returns. Generate a new variable that contains the corresponding log returns (LRET). Calculate the mean, variance, skewness, and kurtosis of the normal and the log returns. Plot the normal against the log returns for MSFT. Briefly discuss your results!

```{r}
PS1_Monthly <- PS1_Monthly %>% 
    mutate(LRET=log(1+RET)) #LRET = ln(1+RET) = ln(Pt/P(t-1))

statistics_monthly <- PS1_Monthly %>% 
    group_by(TICKER) %>% 
#using library(PerformanceAnalytics) to calculate the statistics
    summarise(mean_normal = mean(RET),
              variance_normal = var(RET),
              skewness_normal = skewness(RET),
              kurtosis_normal = kurtosis(RET),
              mean_log = mean(LRET),
              variance_log = var(LRET),
              skewness_log = skewness(LRET), 
              kurtosis_log = kurtosis(LRET))
print(statistics_monthly)

# plot normal against log returns for MSFT
returns_MSFT <- PS1_Monthly %>% 
    filter(TICKER == 'MSFT') %>% 
    ggplot(aes(x=RET,y=LRET)) +
    geom_point()+
    labs(title = "MSFT return comparison", 
       x = "RET", 
       y = "LRET")+
    theme_fivethirtyeight()+
    NULL

returns_MSFT
```

> 6. Go to CANVAS and download the data for PS1 Daily.xlsx. This file contains two worksheets.
HPR Daily contains the daily holding period returns for the six stocks, the S&P 500 Composite Index
and the value-weighted market portfolio (including dividends) from CRSP. Prices Daily contains the
prices for the six stocks and the S&P 500 Composite Index.

```{r importing PS1 Daily data}
PS1_Daily <- read_excel('PS1_Daily.xlsx',skip=1) %>% 
    mutate(DATE = lubridate::ymd(DATE))  # fix `date`

PS1_Daily_longer <-pivot_longer(PS1_Daily, #pivot data from wide to long
                         cols = c(2:7),
                         names_to = "TICKER",
                         values_to = "RET")


```

> 7.  Construct a daily total return index for MSFT and GE stocks and the S&P 500 index and plot them
against each other. Compare your results with the monthly total return indices from above. Are there
any differences? Discuss.

```{r}
PS1_Daily <- PS1_Daily %>% 
#create daily total return index for the MSFT and GE stocks and the S&P 500 index
  mutate(daily_total_return_index_MSFT = MSFT/MSFT[1]) %>%  
  mutate(daily_total_return_index_GE = GE/GE[1]) %>% 
  mutate(daily_total_return_index_SP500 = SPRTRN/SPRTRN[1])

MSFT_GE <- PS1_Daily %>% 
  ggplot(aes(x = daily_total_return_index_MSFT, y = daily_total_return_index_GE)) +
  geom_point()+
  labs(title = "Daily total return index comparison", 
       x = "MSFT", 
       y = "GE")+
  theme_fivethirtyeight()+
    NULL


MSFT_GE

MSFT_SP500 <- PS1_Daily %>% 
  ggplot(aes(x = daily_total_return_index_MSFT, y = daily_total_return_index_SP500)) +
  geom_point()+
  labs(title = "Daily total return index comparison", 
       x = "MSFT", 
       y = "S&P 500")+
  theme_fivethirtyeight()+
  NULL
  
MSFT_SP500

GE_SP500 <- PS1_Daily %>% 
  ggplot(aes(x = daily_total_return_index_GE, y = daily_total_return_index_SP500)) +
  geom_point()+
  labs(title = "Daily total return index comparison", 
       x = "GE", 
       y = "S&P 500")+
  theme_fivethirtyeight()+
  NULL

GE_SP500

```

> 8. As before, the holding period returns are normal returns. Create log returns. Calculate the mean,
variance, skewness, and kurtosis of the normal and log returns. Compare and discuss your results with
the results from monthly frequency

```{r}
# calculation
PS1_Daily_longer <- PS1_Daily_longer %>% 
    mutate(LRET=log(1+RET)) 
statistics_daily <- PS1_Daily_longer %>% 
    group_by(TICKER) %>% 
#library(PerformanceAnalytics)
    summarise(mean_normal = mean(RET),
              variance_normal = var(RET),
              skewness_normal = skewness(RET),
              kurtosis_normal = kurtosis(RET),
              mean_log = mean(LRET),
              variance_log = var(LRET),
              skewness_log = skewness(LRET), 
              kurtosis_log = kurtosis(LRET))

print(statistics_daily)

```

> 9.  Compare the statistical properties of the log holding period return time series both for monthly and
daily returns. Plot a histogram and discuss how the empirical distributions relate to the normal
distribution.

```{r compare statistical properties of log return}

stat_LRET_monthly <- PS1_Monthly %>% 
  ggplot(aes(x = LRET))+
  geom_histogram()+
  facet_wrap(~TICKER)+
  theme_fivethirtyeight()+
  NULL

stat_LRET_monthly

stat_LRET_daily <- PS1_Daily_longer %>% 
  ggplot(aes(x = LRET))+
  geom_histogram()+
  facet_wrap(~TICKER)+
  theme_fivethirtyeight()+
  NULL

stat_LRET_daily

```
From the histograms of log returns of different holding periods, we can see that distributions are skewed to the left. Probably it is due to some large negative return figures. When the holding period increases from a day to a month, the tails of the distributions become heavier. The monthly log holding period return time series distribution is closer to normal distribution than the daily log return. 


> 10.  Pick three stocks and the S&P 500 index (either you can use MSFT, GE and JPM or adapt the code to
pick three random stocks). You will need the holding period returns (both normal and log returns)
and the total return indices you created.

```{r}
Prices_daily <- read_excel("PS1_Daily.xlsx", sheet = 2, skip = 1)
stocks_RET <- PS1_Daily_longer %>% 
  filter(TICKER == "MSFT"|
         TICKER == "GE"|
         TICKER == "JPM") %>% 
  select(1,4,5) 

stocks_RET_wider <- stocks_RET %>%  
  pivot_wider(names_from = "TICKER",
              values_from = "RET") %>% 
  select(-1)

stocks_RET_squared <-stocks_RET %>% 
  mutate(RET_squared = RET^2) %>%  # mutate RET_squared 
  select(-3) %>%  #delete unnecessary column: RET
  pivot_wider(names_from = "TICKER",
              values_from = "RET_squared") %>% 
  select(-1) #delete the date column


    


```

> 11.  Calculate the covariance matrix for the log return series, using both the returns and returns squared.
Discuss your results briefly.

```{r}
stocks_LRET <- PS1_Daily_longer %>% 
  filter(TICKER == "MSFT"|
         TICKER == "GE"|
         TICKER == "JPM") %>% 
  select(1,4,6) 
 
stocks_LRET_wider <- stocks_LRET %>%  
  pivot_wider(names_from = "TICKER",
              values_from = "LRET") %>% 
  select(-1)
 
#print the covariance matrix for the log return series, using the log returns
print(cov(stocks_LRET_wider))
 
stocks_LRET_squared <- stocks_LRET %>% 
  mutate(LRET_squared = LRET^2) %>%  # mutate LRET_squared 
  select(-3) %>%  #delete unnecessary column: LRET
  pivot_wider(names_from = "TICKER",
              values_from = "LRET_squared") %>% 
  select(-1) #delete the date column

#print the covariance matrix for the log return series, using the log returns squared
print(cov(stocks_LRET_squared))




```

> 12. Plot the ACF (autocorrelation function) for prices, returns, returns squared, and absolute returns.
Discuss the results!

```{r}
#Plot the ACF for prices of MSFT, GE, JPM
price_MSFT <- acf(Prices_daily$MSFT, plot = FALSE)
price_MSFT
plot(price_MSFT, main = "ACF of MSFT daily prices") 

price_GE <- acf(Prices_daily$GE, plot = FALSE)
price_GE
plot(price_GE, main = "ACF of GE daily prices") 

price_JPM <- acf(Prices_daily$GE, plot = FALSE)
price_JPM
plot(price_GE, main = "ACF of JPM daily prices") 

#Plot the ACF for log returns of MSFT, GE, JPM
LRET_MSFT <- acf(stocks_LRET_wider$MSFT, plot = FALSE)
LRET_MSFT
plot(LRET_MSFT, main = "ACF of MSFT daily log returns") 

LRET_GE <- acf(stocks_LRET_wider$GE, plot = FALSE)
LRET_GE
plot(LRET_GE, main = "ACF of GE daily log returns") 

LRET_JPM <- acf(stocks_LRET_wider$JPM, plot = FALSE)
LRET_JPM
plot(LRET_JPM, main = "ACF of JPM daily log returns") 

#Plot the ACF for log return squared of MSFT, GE, JPM
LRET_squared_MSFT <- acf(stocks_LRET_squared$MSFT, plot = FALSE)
LRET_squared_MSFT
plot(LRET_squared_MSFT, main = "ACF of MSFT daily log return squared") 

LRET_squared_GE <- acf(stocks_LRET_squared$GE, plot = FALSE)
LRET_squared_GE
plot(LRET_squared_GE, main = "ACF of GE daily log return squared") 

LRET_squared_JPM <- acf(stocks_LRET_squared$JPM, plot = FALSE)
LRET_squared_JPM
plot(LRET_squared_JPM, main = "ACF of JPM daily log return squared") 


#Plot the ACF for normal returns of MSFT, GE, JPM
RET_MSFT <- acf(stocks_RET_wider$MSFT, plot = FALSE)
RET_MSFT
plot(RET_MSFT, main = "ACF of MSFT daily normal returns") 

RET_GE <- acf(stocks_RET_wider$GE, plot = FALSE)
RET_GE
plot(RET_GE, main = "ACF of GE daily normal returns") 

RET_JPM <- acf(stocks_RET_wider$JPM, plot = FALSE)
RET_JPM
plot(RET_JPM, main = "ACF of JPM daily normal returns") 

#Plot the ACF for normal return squared of MSFT, GE, JPM
RET_squared_MSFT <- acf(stocks_RET_squared$MSFT, plot = FALSE)
RET_squared_MSFT
plot(RET_squared_MSFT, main = "ACF of MSFT daily normal return squared") 

RET_squared_GE <- acf(stocks_RET_squared$GE, plot = FALSE)
RET_squared_GE
plot(RET_squared_GE, main = "ACF of GE daily normal return squared") 

RET_squared_JPM <- acf(stocks_RET_squared$JPM, plot = FALSE)
RET_squared_JPM
plot(RET_squared_JPM, main = "ACF of JPM daily normal return squared") 


#Plot the ACF for absolute return of MSFT, GE, JPM

```
